A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-06 doc/popl2013/evaluation.tex

Figures \ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
component of each bar shows the overhead added by the dynamic checks.  In 
each figure the problem size stays the same as the number of processors are increased
(strong scaling).  Figure \ref{fig:amr_overhead} includes multiple problem sizes to show
how overhead is affected by changing problem size (weak scaling).  For cases where there
is an existing implementation to compare against we have included a dotted line indicating
baseline performance.  In a few cases (Figures \ref{fig:fluid_overhead} and 
\ref{fig:amr4096}) checking overhead is the difference
between better and worse performance than the baseline.  For an explanation of overall
performance relative to the baseline implementations please refer to \cite{Legion12}.

\begin{figure}
\begin{center}
\subfigure[4096 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_4096_popl.pdf}
\label{fig:amr4096}
}
\subfigure[8192 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_8192_popl.pdf}
\label{fig:amr8192}
}
\subfigure[16384 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_16384_popl.pdf}
\label{fig:amr16384}
}
\end{center}
\vspace{-2mm}
\caption{Overhead for the AMR application for three different problem sizes.\label{fig:amr_overhead}}
\vspace{-6mm}
\end{figure}

@@

\subsection{Scalability}
\label{subsec:scalability}

>>>>

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show 
the total time spent by all CPUs and GPUs in each phase of the application.  The topmost
component of each bar shows the overhead added by the dynamic checks.  In 
each figure the problem size stays the same as the number of processors are increased
(strong scaling).  Figure~\ref{fig:amr_overhead} includes multiple problem sizes to show
how overhead is affected by changing problem size (weak scaling).  For cases where there
is an existing implementation to compare against we have included a dotted line indicating
baseline performance.  In a few cases (Figures~\ref{fig:fluid_overhead} and 
\ref{fig:amr4096}) checking overhead is the difference
between better and worse performance than the baseline.  For an explanation of overall
performance relative to the baseline implementations please refer to \cite{Legion12}.

\begin{figure}
\begin{center}
\subfigure[4096 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_4096_popl.pdf}
\label{fig:amr4096}
}
\subfigure[8192 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_8192_popl.pdf}
\label{fig:amr8192}
}
\subfigure[16384 cells per dimension]
{
\includegraphics[scale=0.35]{figs/amr_16384_popl.pdf}
\label{fig:amr16384}
}
\end{center}
\vspace{-2mm}
\caption{Overhead for the AMR application.\label{fig:amr_overhead}}
\vspace{-6mm}
\end{figure}

@@

\subsection{Scalability}
\label{subsec:scalability}
The scalability of Legion derives directly from Theorem~\ref{thm:hiersched}.  This 
theorem makes it safe for the runtime to only perform dependence checks between two
tasks that share the same immediate parent task instead of having to consider all
possible interactions with other tasks anywhere in the system.  Since all tasks that
share the same immediate parent are all created on the node where the parent is executing,
all dependence checks and scheduling decisions can be performed locally with no communication.  
Without this property the runtime would be forced to inform all other processors in the system 
of every task that it launches using expensive broadcasts.

Figures~\ref{fig:ckt_overhead}, \ref{fig:fluid_overhead}, and \ref{fig:amr_overhead} show that
the overhead of the Legion runtime is always less than 7\% of the total execution time
of the applications.  In some applications communication doesn't scale well, but in these cases communication is
the result of the algorithm being performed and is not caused by runtime communication.
Even in the case of the Circuit application which exhibits quadratic communication, the Legion
runtime is able to achieve a 62.5X speedup on 96 GPUs over a hand-coded single GPU implementation\cite{Legion12}.

### 2012-07-06 doc/popl2013/intro.tex

Legion programs organize data into {\em logical regions}.  Logical
regions (which we usually refer to simply as {\em regions}) express
locality, which is just data that will be used together and therefore
should be colocated in the machine.  Listing~\ref{lst:linkedlist}
gives an example that defines a region containing a linked list of integers and a task
that sums the elements of the list.  As is standard in region systems, we use
$T @ r$ for the type of a pointer to a value of type $T$ in region $r$.
As is also

>>>>

Legion programs organize data into {\em logical regions}.  Logical
regions (which we usually refer to simply as {\em regions}) express
{\em locality}, which is just data that will be used together and
therefore should be colocated in the machine.
Listing~\ref{lst:linkedlist} gives an example that defines a region
containing a linked list of integers and a task that sums the elements
of the list.  As is standard in region systems, we use $T @ r$ for the
type of a pointer to a value of type $T$ in region $r$.  The task also
illustrates a simple use of the permissions system; in this case the
task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
to read the values in region {\tt r}.

Listing~\ref{lst:treereverse} gives a more elaborate example.  While
artificial (it is designed to use the greatest number of Legion
features in the smallest space), the program illustrates typical
Legion programming idioms.  Given a list of integers as defined in Listing~\ref{lst:linkedlist}, 
the task {\tt make\_tree} builds a binary tree of pointers to list elements.
Each internal node of the tree stores a pointer to a list element with value $i$
and the left and right subtrees hold pointers to list elements with values less than $i$ and greater than $i$,
respectively.  The task {\tt reverse\_tree} then does an in-place, recursive reversal of the children of each interior
node of the tree.  We focus on a description of the structure of types and permissions in this example.

Line 1 defines a type that is a {\em region relationship} {\tt rr[ rll, rlr, rtl, rtr]}.  Region relationships are
constrained existential types.  The {\tt tree} region relationship defines the type of a node of the tree: it has a pointer
to a list node and two pointers to the left and right subtrees. The crucial part is the {\em subregion} and {\em disjointness}
constraints that describe the relationships between the various regions.  Here {\tt rl} is the region holding all list elements,
{\tt rls} is the region holding

### 2012-07-06 doc/popl2013/legion-types.tex
Fixing broken references

\bibliographystyle{plainnat}
\usepackage{amsmath,amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}
\usepackage{multirow}

@@

\begin{abstract}
Legion is a programming model that introduces logical regions as a mechanism for targeting large heterogeneous machines with complex memory hierarchies. Logical regions are first-class values in Legion, allowing programmers to specify independence and locality properties of program data dynamically. We present a type system with permission and coherence properties for reasoning about logical regions.  We prove the soundness of the  type system and use it to show that all memory accesses in Legion are safe. Furthermore, we show that type soundness directly enables a scalable, hierarchical scheduling algorithm for Legion programs. We implement a type checker for Legion programs and improve performance up to 71% by eliding memory checks the type system proves safe. We also demonstrate that an implementation of the Legion runtime based on the hierarchical scheduling algorithm scales well for three real-world applications.

>>>>

\bibliographystyle{plainnat}
\usepackage{amsmath,amsthm}
\usepackage{listings}
\usepackage{tikz}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{cancel}
\usepackage{multirow}
\usepackage{ifmtarg}
\usepackage{url}

@@

\begin{abstract}

Legion is a programming model that introduces 
{\em logical regions} as a mechanism for targeting large heterogeneous machines 
with complex memory hierarchies.  Logical regions are first-class 
values in Legion, allowing programmers to specify independence 
and locality properties of program data dynamically.  We present
a type system with permissions and coherence properties for reasoning about logical regions.  
We prove the soundness of the type system and use it to show
that all memory accesses in Legion are safe.  Furthermore, we prove that
type soundness directly enables a scalable, hierarchical
scheduling algorithm for Legion programs.  We implement a type checker
for Legion programs and improve performance up to 71\% by eliding
memory checks the type system proves safe.  We also demonstrate that an implementation of the Legion 
runtime based on the hierarchical scheduling algorithm scales well for
three real-world applications.
